---
title: "Comprehensive Analysis of Stock Price Time Series: Clustering and Forecasting Techniques"
author: "Tejesh Pawar"
format: revealjs
toc: true
slide-number: true
smaller: false
scrollable: true
---

## Overview
- The goal of this project is to demonstrate time series analysis techniques, including clustering and forecasting, using historical stock price data.
- We'll provide an overview of K-means clustering and time series forecasting and discuss their applications in grouping stocks based on price movement patterns and predicting future stock prices.
---

## Theoretical Explanation
### K-means Clustering
- **Unsupervised Clustering**: K-means clustering is an unsupervised machine learning algorithm used to partition a dataset into K distinct clusters based on similarity.
- **Objective**: The algorithm aims to minimize within-cluster variance, which is the sum of squared distances between data points and their assigned cluster centroids.
- **Algorithm Steps**:
  1. **Initialization**: Randomly select K initial cluster centroids.
  2. **Assignment**: Assign each data point to the nearest centroid.
  3. **Update Centroids**: Recalculate the centroids based on the mean of data points in each cluster.
  4. **Iteration**: Repeat steps 2 and 3 until convergence.
- **Applications**: K-means clustering is used in various fields, including customer segmentation, image segmentation, and anomaly detection.
- **Mathematical Representation**:
  - Objective Function:
    $$
    J(c, \mu) = \sum_{i=1}^{m} \sum_{j=1}^{k} \| x^{(i)} - \mu_j \|^2
    $$
    Where:
    - \(J(c, \mu)\) is the objective function.
    - \(c^{(i)}\) is the cluster assignment for the \(i\)th example.
    - \(\mu_j\) is the centroid of the \(j\)th cluster.
    - \(k\) is the number of clusters.

##

### Time Series Forecasting
- Definition: Exponential smoothing is a popular time series forecasting method that assigns exponentially decreasing weights to past observations.
- Smoothing Parameter(α): Controls the rate of decay of past observations' influence on the forecast.
- Implementation: Requires only historical data and a choice of 
𝛼
- Adaptability: Can be easily adjusted to capture different patterns in data, such as trend and seasonality.
- Applications: Widely used in finance, economics, and operations management for short-term forecasting tasks.

## Sample Dataset
- We'll use historical daily stock price data for a few companies like Amazon, Netflix, Microsoft, Alphabet,Apple and Meta, including columns for the date and closing price.
- Additionally, we'll include features such as volume and technical indicators (e.g., moving averages) for forecasting.

## R Code Implementation
```{r, echo = TRUE}
# Load required libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(forecast)
```

```{r, echo = TRUE}
# Load the sample dataset
data <- read.csv('Tech_data.csv')
head(data, 10)
```

```{r, echo = TRUE}
# Preprocess the data (if needed)
data$time <- as.Date(data$time)  # Convert 'time' column to Date format
```

```{r, echo = TRUE}
# Select features for clustering (e.g., closing price)
features <- dplyr::select(data, open, high, low, close)
```

```{r, echo = TRUE}
```


```{r, echo = TRUE}
# Standardize the features
scaled_features <- scale(features)
```

```{r, echo = TRUE}
# Apply K-means clustering
k <- 3  # Number of clusters
kmeans_result <- kmeans(scaled_features, centers=k)
```

```{r, echo = TRUE}
# Add cluster labels to the dataset
data$Cluster <- as.factor(kmeans_result$cluster)
```

## Visualize the clustering result
```{r, echo = TRUE}
# Visualize the clustering result
ggplot(data, aes(x = time, y = close, color = Cluster)) +
  geom_line() +
  labs(title = "K-means Clustering of Stock Price Time Series", x = "Time", y = "Close Price") +
  facet_wrap(~Company, scales = "free_y")
```

#
```{r, echo = TRUE}
# Time Series Forecasting (Exponential Smoothing)
# Convert 'time' column to Date format
data$time <- as.Date(data$time)
```

```{r, echo = TRUE}
# Prepare time series data
ts_data <- ts(data$close, frequency = 1) 
```

```{r, echo = TRUE}
# Split data into training and test sets
train_end <- as.Date("1999-11-01")  # Training data until November 1999
test_start <- as.Date("1999-12-01")  # Test data from December 1999 onwards

train <- window(ts_data, end = as.numeric(train_end))  
test <- window(ts_data, start = as.numeric(test_start))
```

```{r, echo = TRUE}
# Perform exponential smoothing
forecast_model <- ets(train)
```

## Generate forecasts
```{r, echo = TRUE}
# Generate forecasts
forecast_values <- forecast(forecast_model, h=length(test))
forecast_values
```

## Plot forecasts
```{r, echo = TRUE}
# Plot forecasts
autoplot(forecast_values) + 
  autolayer(fitted(forecast_model), series = "Predicted") +
  autolayer(test, series = "Actual") +
  xlab("Time") + ylab("Close Price") +
  ggtitle("Exponential Smoothing Forecasting")
```

# Conclusion

- Explored K-means clustering and time series forecasting for stock price analysis.
- K-means clusters similar stocks to identify patterns in price movements.
- Exponential smoothing forecasts future stock prices, aiding investment decisions.
- Integration of these techniques enhances understanding and informs investment strategies.

# Thank You!
